---
title: "Coming from 'tidyverse'"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Coming from 'tidyverse'}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = "")
knitr::opts_chunk$set(
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 300
)

pkgs <- c(
  "datawizard",
  "tidyr",
  "dplyr",
  "see",
  "ggplot2",
  "parameters",
  "lme4"
)

if (!all(sapply(pkgs, requireNamespace, quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
}
```


```{css, echo=FALSE, eval = TRUE}
.datawizard, .datawizard > .sourceCode {
  background-color: #e6e6ff;
}
.tidyverse, .tidyverse > .sourceCode {
  background-color: #d9f2e5;
}
```

# Introduction

`{datawizard}` package aims to make basic data wrangling easier than 
with base R. The data wrangling workflow it supports is similar to the one
supported by the tidyverse package combination of `{dplyr}` and `{tidyr}`. However,
one of its main features is that it has a very few dependencies: `{stats}` and `{utils}`
(included in base R) and `{insight}`, which is the core package of the _easystats_ 
ecosystem. This package grew organically to simultaneously satisfy the 
"0 non-base hard dependency" principle of _easystats_ and the data wrangling needs
of the constituent packages in this ecosystem.

One drawback of this genesis is that not all features of the `{tidyverse}` 
packages are supported since only features that were necessary for _easystats_ 
ecosystem have been implemented. Some of these missing features (such as `summarize`
or the pipe operator `%>%`) are made available in other dependency-free packages, 
such as [`{poorman}`](https://github.com/nathaneastwood/poorman/). It is also 
important to note that `{datawizard}` was designed to avoid namespace collisions 
with `{tidyverse}` packages.

In this article, we will see how to go through basic data wrangling steps with 
`{datawizard}`. We will also compare it to the `{tidyverse}` syntax for achieving the same. 
This way, if you decide to make the switch, you can easily find the translations here.
This vignette is largely inspired from `{dplyr}`'s [Getting started vignette](https://dplyr.tidyverse.org/articles/dplyr.html).

```{r eval = TRUE}
library(dplyr)
library(tidyr)
library(datawizard)
```

# Workhorses

Before we look at their *tidyverse* equivalents, we can first have a look at 
`{datawizard}`'s key functions for data wrangling:

| Function          | Operation                                         |
| :---------------- | :------------------------------------------------ |
| `data_filter()`   | [to select only certain observations](#filtering) |
| `data_select()`   | [to select only a few variables](#selecting)      |
| `data_arrange()`  | [to sort observations](#sorting)      |
| `data_extract()`  | [to extract a single variable](#extracting)       |
| `data_rename()`   | [to rename variables](#renaming)                  |
| `data_relocate()` | [to reorder a data frame](#relocating)            |
| `data_to_long()`  | [to convert data from wide to long](#reshaping)   |
| `data_to_wide()`  | [to convert data from long to wide](#reshaping)   |
| `data_join()`     | [to join two data frames](#joining)               |

Note that there are a few functions in `{datawizard}` that have no strict equivalent
in `{dplyr}` or `{tidyr}` (e.g `data_rotate()`), and so we won't discuss them in
the next section.

# Equivalence with `{dplyr}` / `{tidyr}`

Before we look at them individually, let's first have a look at the summary table of this equivalence.

| Function          | Tidyverse equivalent(s)                                             |
| :---------------- | :------------------------------------------------------------------ |
| `data_filter()`   | `dplyr::filter()`, `dplyr::slice()`                                 |
| `data_select()`   | `dplyr::select()`                                                   |
| `data_arrange()`  | `dplyr::arrange()`                                                  |
| `data_extract()`  | `dplyr::pull()`                                                     |
| `data_rename()`   | `dplyr::rename()`                                                   |
| `data_relocate()` | `dplyr::relocate()`                                                 |
| `data_to_long()`  | `tidyr::pivot_longer()`                                             |
| `data_to_wide()`  | `tidyr::pivot_wider()`                                              |
| `data_join()`     | `dplyr::inner_join()`, `dplyr::left_join()`, `dplyr::right_join()`, |
|                   | `dplyr::full_join()`, `dplyr::anti_join()`, `dplyr::semi_join()`    |
| `data_peek()`     | `dplyr::glimpse()`                                                  |

## Filtering {#filtering}

`data_filter()` is a wrapper around `subset()`. Therefore, if you want to have
several filtering conditions, you need to use `&`. Separating the conditions
with a comma (as in `dplyr::filter()`) will **not** work; it will only apply the
first condition.

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r filter, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_filter(skin_color == "light" &
    eye_color == "brown")
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  filter(
    skin_color == "light",
    eye_color == "brown"
  )
```
:::

::::

```{r filter, eval = TRUE, echo = FALSE}
```


<!-- Shorten output to make it easier to read: -->
```{r, echo = FALSE, eval = TRUE}
starwars <- head(starwars)
```

## Selecting {#selecting}

`data_select()` is the equivalent of `dplyr::select()`. 
The main difference between these two functions is that `data_select()` uses two
arguments (`select` and `exclude`) and requires quoted column names if we want to 
select several variables, while `dplyr::select()` accepts any unquoted column names.

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r select1, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_select(select = c("hair_color", "skin_color", "eye_color"))
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  select(hair_color, skin_color, eye_color)
```
:::

::::

```{r select1, eval = TRUE, echo = FALSE}
```

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r select2, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_select(select = -ends_with("color"))
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  select(-ends_with("color"))
```
:::

::::

```{r select2, eval = TRUE, echo = FALSE}
```

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

<!-- TODO: Although we say the column names need to be quoted, they are unquoted
here and quoting them won't work. Should we comment on that? -->

```{r select3, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_select(select = -hair_color:eye_color)
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  select(!(hair_color:eye_color))
```
:::

::::

```{r select3, eval = TRUE, echo = FALSE}
```


:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r select4, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_select(exclude = regex("color$"))
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  select(-contains("color$"))
```
:::

::::

```{r select4, eval = TRUE, echo = FALSE}
```


:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r select5, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_select(select = is.numeric)
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  select(where(is.numeric))
```
:::

::::

```{r select5, eval = TRUE, echo = FALSE}
```

You can find a list of all the select helpers with `?data_select`.



## Sorting {#sorting}

`data_arrange()` is the equivalent of `dplyr::arrange()`. It takes two arguments:
a data frame, and a vector of column names used to sort the rows. Note that contrary
to most other functions in `{datawizard}`, it is not possible to use select helpers
such as `starts_with()` in `data_arrange()`.

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}
:::{}
```{r arrange1, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_arrange(c("hair_color", "height"))
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  arrange(hair_color, height)
```
:::

::::

```{r arrange1, eval = TRUE, echo = FALSE}
```

You can also sort variables in descending order by putting a `"-"` in front of 
their name, like below:

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}
:::{}
```{r arrange2, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_arrange(c("-hair_color", "-height"))
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  arrange(desc(hair_color), -height)
```
:::

::::

```{r arrange2, eval = TRUE, echo = FALSE}
```


## Extracting {#extracting}

Although we mostly work on data frames, it is sometimes useful to extract a single 
column as a vector. This can be done with `data_extract()`, which reproduces the 
behavior of `dplyr::pull()`:

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}
:::{}
```{r extract1, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_extract(gender)
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  pull(gender)
```
:::

::::

```{r extract1, eval = TRUE, echo = FALSE}
```

We can also specify several variables in `select`. In this case, `data_extract()`
is equivalent to `data_select()`:

```{r eval = TRUE}
starwars %>%
  data_extract(select = contains("color"))
```




## Renaming {#renaming}

`data_rename()` is the equivalent of `dplyr::rename()` but the syntax between the 
two is different. While `dplyr::rename()` takes new-old pairs of column
names, `data_rename()` requires a vector of column names to rename, and then 
a vector of new names for these columns that must be of the same length.

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r rename1, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_rename(
    pattern = c("sex", "hair_color"),
    replacement = c("Sex", "Hair Color")
  )
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  rename(
    Sex = sex,
    "Hair Color" = hair_color
  )
```
:::

::::

```{r rename1, eval = TRUE, echo = FALSE}
```

The way `data_rename()` is designed makes it easy to apply the same modifications 
to a vector of column names. For example, we can remove underscores and use 
TitleCase with the following code:

```{r rename2}
to_rename <- names(starwars)

starwars %>%
  data_rename(
    pattern = to_rename,
    replacement = tools::toTitleCase(gsub("_", " ", to_rename))
  )
```

```{r rename2, eval = TRUE, echo = FALSE}
```

It is also possible to add a prefix or a suffix to all or a subset of variables 
with `data_addprefix()` and `data_addsuffix()`. The argument `select` accepts 
all select helpers that we saw above with `data_select()`:

```{r rename3}
starwars %>%
  data_addprefix(
    pattern = "OLD.",
    select = contains("color")
  ) %>%
  data_addsuffix(
    pattern = ".NEW",
    select = -contains("color")
  )
```

```{r rename3, eval = TRUE, echo = FALSE}
```

## Relocating {#relocating}

Sometimes, we want to relocate one or a small subset of columns in the dataset.
Rather than typing many names in `data_select()`, we can use `data_relocate()`,
which is the equivalent of `dplyr::relocate()`. Just like `data_select()`, we can
specify a list of variables we want to relocate with `select` and `exclude`.
Then, the arguments `before` and `after`^[Note that we use `before` and `after` 
whereas `dplyr::relocate()` uses `.before` and `.after`.] specify where the selected columns should
be relocated:

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r relocate1, class.source = "datawizard"}
# ---------- datawizard -----------
starwars %>%
  data_relocate(sex:homeworld, before = "height")
```
:::
  
::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
starwars %>%
  relocate(sex:homeworld, .before = height)
```
:::
  
::::

```{r relocate1, eval = TRUE, echo = FALSE}
```

In addition to column names, `before` and `after` accept column indices. Finally,
one can use `before = -1` to relocate the selected columns just before the last 
column, or `after = -1` to relocate them after the last column.

```{r eval = TRUE}
# ---------- datawizard -----------
starwars %>%
  data_relocate(sex:homeworld, after = -1)
```


## Reshaping {#reshaping}

### Longer

Reshaping data from wide to long or from long to wide format can be done with
`data_to_long()` and `data_to_wide()`. These functions were designed to match 
`tidyr::pivot_longer()` and `tidyr::pivot_wider()` arguments, so that the only 
thing to do is to change the function name. However, not all of 
`tidyr::pivot_longer()` and `tidyr::pivot_wider()` features are available yet. 

We will use the `relig_income` dataset, as in the [`{tidyr}` vignette](https://tidyr.tidyverse.org/articles/pivot.html).

```{r eval = TRUE}
relig_income
```


We would like to reshape this dataset to have 3 columns: religion, count, and 
income. The column "religion" doesn't need to change, so we exclude it with 
`-religion`. Then, each remaining column corresponds to an income category. 
Therefore, we want to move all these column names to a single column called 
"income". Finally, the values corresponding to each of these columns will be 
reshaped to be in a single new column, called "count".

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r pivot1, class.source = "datawizard"}
# ---------- datawizard -----------
relig_income %>%
  data_to_long(
    -religion,
    names_to = "income",
    values_to = "count"
  )
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
relig_income %>%
  pivot_longer(
    !religion,
    names_to = "income",
    values_to = "count"
  )
```
:::

::::

```{r pivot1, eval = TRUE, echo = FALSE}
```


To explore a bit more the arguments of `data_to_long()`, we will use another
dataset: the `billboard` dataset.
```{r eval = TRUE}
billboard
```

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r pivot2, class.source = "datawizard"}
# ---------- datawizard -----------
billboard %>%
  data_to_long(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  )
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
billboard %>%
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  )
```
:::

::::

```{r pivot2, eval = TRUE, echo = FALSE}
```


### Wider

Once again, we use an example in the `{tidyr}` vignette to show how close `data_to_wide()`
and `pivot_wider()` are:
```{r eval=TRUE}
fish_encounters
```


:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r pivot3, class.source = "datawizard"}
# ---------- datawizard -----------
fish_encounters %>%
  data_to_wide(
    names_from = "station",
    values_from = "seen",
    values_fill = 0
  )
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
fish_encounters %>%
  pivot_wider(
    names_from = station,
    values_from = seen,
    values_fill = 0
  )
```
:::

::::

```{r pivot3, eval = TRUE, echo = FALSE}
```



## Joining {#joining}

<!-- explain a bit more the args of data_join -->

In `{datawizard}`, joining datasets is done with `data_join()` (or its alias 
`data_merge()`). Contrary to `{dplyr}`, this unique function takes care of all 
types of join, which are then specified inside the function with the argument
`join` (by default, `join = "left"`).

Below, we show how to perform the four most common joins: full, left, right and 
inner. We will use the datasets `band_members`and `band_instruments` provided by `{dplyr}`:

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r eval = TRUE}
band_members
```
:::

::: {}

```{r eval = TRUE}
band_instruments
```
:::

::::


### Full join

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r join1, class.source = "datawizard"}
# ---------- datawizard -----------
band_members %>%
  data_join(band_instruments, join = "full")
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
band_members %>%
  full_join(band_instruments)
```
:::

::::

```{r join1, eval = TRUE, echo = FALSE}
```



### Left and right joins

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r join2, class.source = "datawizard"}
# ---------- datawizard -----------
band_members %>%
  data_join(band_instruments, join = "left")
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
band_members %>%
  left_join(band_instruments)
```
:::

::::

```{r join2, eval = TRUE, echo = FALSE}
```


:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r join3, class.source = "datawizard"}
# ---------- datawizard -----------
band_members %>%
  data_join(band_instruments, join = "right")
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
band_members %>%
  right_join(band_instruments)
```
:::

::::

```{r join3, eval = TRUE, echo = FALSE}
```



### Inner join

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r join4, class.source = "datawizard"}
# ---------- datawizard -----------
band_members %>%
  data_join(band_instruments, join = "inner")
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
band_members %>%
  inner_join(band_instruments)
```
:::

::::

```{r join4, eval = TRUE, echo = FALSE}
```



# Other useful functions

`{datawizard}` contains other functions that are not necessarily included in 
`{dplyr}` or `{tidyr}` or do not directly modify the data. Some of them are 
inspired from the package `janitor`. 

## Work with rownames

We can convert a column in rownames and move rownames to a new column with
`rownames_as_column()` and `column_as_rownames()`:

```{r eval = TRUE}
mtcars <- head(mtcars)
mtcars

mtcars2 <- mtcars %>%
  rownames_as_column(var = "model")

mtcars2

mtcars2 %>%
  column_as_rownames(var = "model")
```


## Work with column names

When dealing with messy data, it is sometimes useful to use a row as column
names, and vice versa. This can be done with `row_to_colnames()` and
`colnames_to_row()`.

```{r eval = TRUE}
x <- data.frame(
  X_1 = c(NA, "Title", 1:3),
  X_2 = c(NA, "Title2", 4:6)
)
x
x2 <- x %>%
  row_to_colnames(row = 2)
x2

x2 %>%
  colnames_to_row()
```

## Take a quick look at the data

:::: {style="display: grid; grid-template-columns: 50% 50%; grid-column-gap: 10px;"}

::: {}

```{r glimpse, class.source = "datawizard"}
# ---------- datawizard -----------
data_peek(iris)
```
:::

::: {}

```{r, class.source = "tidyverse"}
# ---------- tidyverse -----------
glimpse(iris)
```
:::

::::

```{r glimpse, eval = TRUE, echo = FALSE}
```
